---
title: "Clustering by Zip Codes in the US"
author: "Junyan Huang"
date: "October 7, 2018"
output: word_document
---

#Introduction
This is project is to find the top 10 largest cities in the US by applying the hierarchical clustering techniques on the zip codes.

### My Method
Due to the large size of the dataset, it is very slow to run the hierarchical clustering on my PC. Thus, I randomly sample the data into a smaller one and use the sample to do clustering. 

The hierarchical clustering ends by conbining all observations into one cluster. In order to get the cities, I compute the number of unique city names in the sample data, and then use the cutree function to divide the dendrogram tree into such number of clusters. The clusters will not be 100% accurate. So I compute the most frequent city in each cluster and take it as the clustered city.

#Import, clean and prepare the dataset.
R has a zipcode package that holds the zip code dataset.
```{r}
library(zipcode)
data(zipcode)
#head(zipcode)
```

Check if the dataset has any missing values.
```{r}
empty <- zipcode[!complete.cases(zipcode),]
#print(empty)
```
There are 647 zip codes missing latitude and longitude. 

Remove those rows with missing values.
```{r}
zipcode_no_missing<- zipcode[complete.cases(zipcode),]
```

```{r}
summary(zipcode_no_missing)
```

Plot the US map based on the zip codes coordinates. The map will show density.
```{r}
library(ggplot2)
ggplot(zipcode_no_missing, aes(longitude, latitude)) +xlim(-180, 180)+ylim(-45, 75)+stat_density2d(aes(alpha=..level..), geom="polygon") +
  scale_alpha_continuous(limits=c(0,0.5),breaks=seq(0,0.2,by=0.025))+
  geom_point(colour=139,alpha=0.05)+
  theme_bw()
```

The summary  and the plot indicate that the US is ranged from -44.25 to 71.30 in latitude and - 176.64 to 171.18 in longitude. But base on Wikipedia, the US Lat-long coorditates are: latitude from 19.50139 to 64.85694 and longitude from -161.75583 to -68.01197. Those outliers should be places beyond the mianland. 

Remove outliers in the dataset (also remove Alaska and Hawaii since they are obviously not large cities.)
```{r}
lat_low = 20
lat_up = 51
long_low=-150
long_upper = -50
mydata1<- subset(zipcode_no_missing, latitude>=lat_low&latitude<=lat_up)
mydata<- subset(mydata1, longitude<=long_upper&longitude>=long_low)
```

Here is the map that contains only the mainland and shows the density by color.
```{r}
ggplot(mydata, aes(longitude, latitude))+stat_density2d(aes(alpha=..level..), geom="polygon") +
  scale_alpha_continuous(limits=c(0,0.5),breaks=seq(0,0.2,by=0.025))+
  geom_point(colour=139,alpha=0.05)+
  theme_bw()
```

From the map above, we can see the regions with high density are the eastern half and the west coast. The majority of the western part is a low-density region.

## Randomly sample the dataset.
The dataset is very large. It is very slow to run the hierarchical clustering. So I randomly sample the dataset. 

```{r}
set.seed(123)
n <- nrow(mydata)
index<-sample(1:n, 5000)
mysample<- mydata[index,]
```

```{r}
ggplot(mysample, aes(longitude, latitude))+stat_density2d(aes(alpha=..level..), geom="polygon") +
  scale_alpha_continuous(limits=c(0,0.5),breaks=seq(0,0.2,by=0.025))+
  geom_point(colour=139,alpha=0.05)+
  theme_bw()
```

The sample data still tells the relatively correct density of the mainland. So it should be a good sample.


```{r}
summary(mysample)
dim(mysample)
```


# Cluster analysis on the zip codes.
I will use hierarchical clustering.

```{r}
library(ClustGeo)
```

```{r}
D <- dist(mysample[, 4:5])
tree <- hclust(D)
```

```{r}
sum(tree$height)
```

```{r}
city_number=length(unique(mysample$city))
city_number
```


```{r}
cluster_assignments <- cutree(tree, k=city_number)
```

```{r}
library(dplyr)
```


```{r}
zipcode_clustered <- mutate(mysample, cluster=cluster_assignments)
```

Filter the top 10 cities based on the number of points in each cluster. The more points in the cluster, the larger the city.
```{r}
#assign the cluster number to the mysample dataset.
mysample$cluster_assigned <- zipcode_clustered$cluster
top10_clusters <- names(head((sort(table(mysample$cluster_assigned), decreasing = TRUE)), 10))
top10_clusters
```

The top 10 clusters with most points are as above.

```{r}
#get the corresponding cities.
top1 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[1]])))
top2 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[2]])))
top3 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[3]])))
top4 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[4]])))
top5 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[5]])))
top6 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[6]])))
top7 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[7]])))
top8 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[8]])))
top9 <- names(which.max(table(mysample$city[mysample$cluster==top10_clusters[9]])))
top10 <- names(which.max(table(mysample$city
                               [mysample$cluster==top10_clusters[10]])))
citynames <- c(top1, top2, top3, top4, top5, top6, top7, top8, top9, top10)
print(citynames)

```

Plot the top 10 cities.
```{r}
top10_cities  <- mysample %>% filter(cluster_assigned%in%top10_clusters)
noise  <- mysample %>% filter(!(cluster_assigned%in%top10_clusters))

ggplot(mysample, aes(x = longitude, y = latitude, alpha = 0.5)) + 
  geom_point(aes(),noise) +
  geom_point(aes(color = as.factor(cluster_assigned)), top10_cities,
             size = 4) +
  coord_map() + 
  theme(legend.position = "none")

```

###The top 10 largest cities:

1. Los Angeles 
2. Washington 
3. Dallas 
4. El Paso
5. New York
6. Houston
7. Salt Lake City
8. Detroit
9. Pittsburgh
10. Boston




